Anomaly detection with kernel density estimation on manifolds

### 2022-02-27 paper/paper.tex
Typo

\begin{document}
\maketitle
\begin{abstract}
Kernel denstity estimation based on random samples often invovles a crutial bandwidth selection. For data points lying on the Riemannien manifolds embedded in a high-dimensional space, manifold learning aims to find the low-dimensional representation of the data but there are local distortion for each observation, which failed to be captured by the kernel density estimation. We propose a new kernel density estimation of manifold learning embedding with the Riemannian metric being the pointwise bandwidth matrix and further identify outliers with lowest densities. The Riemannian metric measures the direction and angle of the local distortion induced by the non-linear functions in manifold learning and is used for smoothing the density estimates. To compare our proposed method with fixed bandwidth KDE, we run two simulation with a 2-D meta data mapped into a 3-D swiss roll or twin peaks shape and a 5-D semi-hypersphere meta data mapped in 100-D space, and show that the proposed estimator could improve the density estimates given a good manifold learning embedding and gives higher correlations between ranks of true density and estimated density. For application, we use the Irish smart meter data and explore the distributions of different households and time periods of the week. Both density estimators are compared by examing the distributions of the most typical households with highest densities and the most anomalous households with lowest densities.
\end{abstract}
\begin{keywords}
manifold learning, variable bandwidth, Riemannian metric, highest density region, Gaussian kernels
\end{keywords}

@@

Anomaly detection has been an important and diverse area where anomalies or outliers are detected in a given data. It often involves a data analysis process to uncover the unusual patterns and has been widely applied to machine learning \autocite{Omar2013-lz}, network intrusions identification \autocite{Ahmed2016-co,Bhuyan2013-ra}, medical imaging \autocite{Fernando2022-ef}, fraudulent attacks \autocite{Ahmed2016-dz}, cyber-security \autocite{Ten2011-co}, energy consumption \autocite{Cheng2021-ex}, the last of which explored detecting households with anomalous electricity usage distributions instead of raw data.
In the case of non-Euclidean sample space, the observations lie on a Riemannian manifold embedded in a very high-dimensional ambient space, which make it computationally expensive or impossible to detect anomalies. To address this problem, we propose a kernel density estimator of the low-dimensional statistical manifold embedding and find outliers as the data with the lowest density estimates.

Kernel density estimation {[}KDE; \textcite{Parzen1962-gt}; \textcite{Chen2017-dw}{]} is one of the most popular method to calculate the probability density function from a sample dataset. KDE is flexible to learn the shape of the underlying density from the data controlled by the bandwidth and the selection of bandwidth is crucial in KDE \autocite{Jones1990-oe,Terrell1992-ut}. There are extensive research on bandwidth selection, two main categories of which are the cross-validation \autocite{Jones1992-ta,Sain1994-gr} and plug-in methods \autocite{Wand1994-gc,Duong2003-sp}. For multivariate data, variable kernel density estimator {[}VKDE; \textcite{Jones1990-oe}; \textcite{Terrell1992-ut}{]} has been studied with an adaptive bandwidth matrix to control the amount of smoothing on the location of the estimated point {[}balloon estimator; \textcite{Terrell1992-ef}{]} or the data point {[}sample smoothing estimator; \textcite{Terrell1992-ef}{]}.
However, these density estimators are based on random samples in the Euclidean space.

For samples points lying on a manifold with the differentiable structure called the Riemannian manifold, \textcite{Pelletier2005-vu} proposed the a kernel density estimator based on the kernel weight depending on the distance between the estimated points and the sample data points. The idea of the estimator is to use a strictly positive function of the geodesic distance on the manifold and then normalize it with the volume density function of the Riemannian manifold for curvature. \autocite{Henry2009-ll} In many application scenarios, we tend to find that the sample points on a manifold are embedded in a much higher-dimensional space, and the kernel density estimator from \textcite{Pelletier2005-vu} is not directly applicable. This is when we introduce manifold learning to reduce the input data dimension.
For these high-dimensional data set, various manifold learning algorithms including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP (see details of these algorithms in \textcite{Cheng2021-ex}), are applied to get a low-dimensional embedding. We propose a kernel density estimator to be applied to the low-dimensional embedding and define outliers as those with lowest densities.

In manifold learning, the underlying idea is that the data lies on a low-dimensional smooth manifold which is embedded in a high-dimensional space. One of the fundamental objectives of manifold learning is to explore the geometry of the dataset, including the distances between points and volumes of regions of data. These intrinsic geometric attributes of the data, such as distances, angles, and areas, however, can be distorted in the low-dimensional embedding, leading to failure to recover the geometry of the manifold \autocite{Goldberg2008-co}. To tackle this problem and measure the distortion incurred in manifold learning, \textcite{Perrault-Joncas2013-pq} propose the Learn metric algorithm to augment any existing embedding output with geometric information in the Riemannian metric of the manifold itself. By applying this algorithm, the outputs of different manifold learning methods can be unified and compared under the same framework, which would highly benefit in improving the effectiveness of the embedding.

The Riemannian metric using the method of \textcite{Perrault-Joncas2013-pq} gives some idea of the distortion of an embedding. Mapping the points through a non-linear function ``stretches'' some regions of space and ``shrinks'' others. The Riemannian gives us an idea of the direction and angle of this stretching, which is informative for learning the manifold.
In variable kernel density estimate, the bandwidth matrix \(H\) is also defined to control the amount of smoothing for each data point.
Therefore, we propose to the use Riemannian metric as the bandwidth matrix in , we could further get the kernel density estimation of the manifold \(\mathcal{M}\). This kernel density estimate can then be used to produce the highest density region plots \autocite{Hyndman1996-lk} for outlier visualization.

The rest of the paper is organized as follows.
In \autoref{vkde}, we present the proposed algorithm to detect anomalies based on variable kernel density estimates of manifold embeddings. In this section, we provide justification for the use of Riemannian metric as the bandwidth of variable kernel density estimation, including the comparison with fixed bandwidth.
\autoref{simulation} is composed with two simulations with the proposed algorithm; the first deals with a 2-dimensional meta data mapped into a 3-D swiss roll or twin peaks data and the second with a 5-D semi-hypersphere mapped in a 100-D space.
\autoref{application} contains the application to visualize and identify anomalies in the Irish smart meter dataset.
Conclusions and discussions are presented in \autoref{conclusion}.

@@

In this section, we introduce the proposed method to detect anomalies based on the kernel density estimates of manifold learning embeddings where the Riemannian matrix is used as the pointwise variable bandwidth to measure the direction and angle of the distortion of the low-dimensional embeddings. \textcite{Perrault-Joncas2013-pq} gives us an idea of how to measure the direction and angle of the distortion using the Riemannian metric and the Riemannian metric is a positive definite square matrix for each data point. To learn the distribution of the low-dimensional embedding, we use the kernel density estimation with the bandwidth matrix being the Riemannian metric. The outliers could then be defined as the points with lowest density estimates. The proposed schematic is shown in Figure \ref{fig:vkde}. The highlighted two steps in Figure \ref{fig:vkde} are the main contributions of this main chapter, replacing the bandwidth matrix \(\pmb{h}_i\) with the Riemannian metric \(\pmb{g}_i\) for each point in variable kernel density estimate, and computing the highest density region plots based on the density estimates, \(\hat{f}(\pmb{x})\), for anomaly detection.

@@

In general, a multivariate variable kernel density estimate of \(f\) at point \(\pmb{x} \in \mathcal{R}^d\) is defined as
\[
\hat{f}(\pmb{x})=\sum\limits_{i=1}^N K_{\pmb{h}_i}(\pmb{x}-\pmb{a}_i),
\]
where \(a_i, i = 1, 2, \dots, N\) is the random sample drawn from a density \(f\), \(\pmb{h}_i\) is a \(d\times d\) symmetric positive definite matrix that is variable at \(i\), and \(K_{\pmb{h}_i}\) is the kernel function
\[
K_{\pmb{h}_i}(\pmb{x}-\pmb{a}_i)=(2\pi)^{-d/2}|\pmb{h}|^{-1/2}\exp\left[-\frac{1}{2}(\pmb{x}-\pmb{a}_i)'\pmb{h}_i^{-1}(\pmb{x}-\pmb{a}_i)\right],
\]
with \(|\pmb{h}_i|\) being the determinant of matrix \(\pmb{h}_i\). The bandwidth matrix is important in controlling the smoothing across all data points and the shape is determined by the kernel function \(K(\pmb{x})\). We use the sample smooth estimator for each data point, which make a kernel density estimate with local smoothing.
\textcite{Duong2007-up} present an R package \emph{ks} that implement kernel density estimates with fixed bandwidth and variable bandwidth, but it is limited to bandwidth vector or diagonal matrix and is not applicable to full bandwidth matrix.

@@

The Riemannian metric \(g\) is a symmetric and positive definite tensor field which defines an inner product \(<,>_g\) on the tangent space \(T_p\mathcal{M}\) for every point \(p \in \mathcal{M}\).
If the inner product of the tangent space is known for a given geometry, the Riemannian metric is a good measure to recover the geometry of manifold.
The Metric Learning algorithm \autocite{Perrault-Joncas2013-pq} then augment the embedded manifold with the Riemannian metric and produce a Riemannian manifold \((\mathcal{M}, g)\).

@@

As pointed out by \textcite{Perrault-Joncas2013-pq}, if the embedding dimension \(s\) is larger than the manifold intrinsic dimension \(d\), the rank of the embedding metric \(h_n(p)\) is \(d\); otherwise, the Riemannian metric \(g_n\) will be returned.
This algorithm is also implemented in a Python library \emph{megaman} \autocite{McQueen2016-xz}. It is designed to apply the manifold learning methods to large-scale data sets, as well as computing the Riemannian metric of the manifold.

@@

If the embedding dimension \(d=2\), the HDR plots \autocite{Hyndman1996-lk} could be used to identify the relative location of outliers in the embedding and find which observations lie in a highest density region of specified coverage, eg. 1\%, 50\%, 99\%, \textgreater99\%.

@@

In this section, we exam two scenarios for both low and high dimensions to test our proposed algorithm. For visualization purpose, \autoref{twodgaussian} presents a 2-D meta data example. We first simulate the data of size \(N=2,000\) from a mixture of four Gaussian kernels with the same covariance but different means, each consisting of \(500\) points. Different mapping functions are then applied to the 2-D meta data to be mapped in a 3-D feature space, which gives the higher-dimensional input for different manifold learning algorithms, including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP. The embedded dimension is set as \(d=2\), same as the meta data dimension. This enables us to compare the manifold learning embedding with the true meta data. We could now apply Algorithm \ref{alg:vkderm} to get the density estimates of all data points and further detect anomalies. As a high-dimensional example, the second simulation in \autoref{fivedgaussian} is based on a 5-D meta data of size \(N=2,000\) embedded in a 100-D space and the corresponding embedding dimension is \(d=5\).

@@

Given the 2-D meta data, multiple mapping functions could be applied to embed the data in a 3-D space. One of the most famous example in manifold learning is the swiss roll data, with the mapping function in \eqref{eq:swissroll}. The two-dimensional meta data \((\pmb{X}_1, \pmb{X}_2)^\prime\) is transformed into the three-dimensional data \((\pmb{X}, \pmb{Y}, \pmb{Z})^\prime\), shown in the left plot of \autoref{fig:mappings}. The four colors in the mappings represents the four Gaussian kernels used to generate the meta data \((\pmb{X}_1, \pmb{X}_2)^\prime\).
\begin{equation}
\label{eq:swissroll}
\left\{
\begin{array}{lcl}
X = X_1 \cos{X_1}, \\
Y = X_2, \\
Z = X_1 \sin{X_1}.
\end{array}
\right.
\end{equation}

@@

\caption{Highest density region plots of five manifold learning embeddings of the swiss roll data. Colors are indicating densities from left: true densities from the Gaussian mixture model; middle: KDE with Riemannian matrix as variable bandwidth; and right: KDE with fixed bandwidth. Variable KDE preforms better in finding kernel structures with ISOMAP, LLE, and Laplacian Eigenmaps, and in locating outliers with ISOMAP and LLE. The t-SNE and UMAP embeddings are highly distorted and the outliers found are clustered.}\label{fig:sroutliers}
\end{figure}

Now we are able to apply different manifold learning algorithms to \((\pmb{X}, \pmb{Y}, \pmb{Z})^\prime\) and reduce the dimension back to \(d=2\), and further estimate the density of the 2-D embedding. According to the density estimates, we could rank the data points and then identify which observations lie in a highest density region of specified coverage, eg. 1\%, 5\%, 50\%, 99\%, \textgreater99\%. For each of the five manifold learning methods, namely ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, \autoref{fig:sroutliers} presents the 2-D embedding plot in the same row, with the colors indicating the densities levels, the left column for true densities from the Gaussian mixture model, the middle column for highest density region plots with densities from our proposed variable KDE method, and the right for similar HDR plots with densities from KDE with fixed bandwidth. The top twenty outliers with lowest densities are highlighted in black with point indexes in blue.
From \autoref{fig:metadensity} and the data generating process, we know that there are four highest density regions. However, in all manifold learning embeddings colored with true densities (left column in \autoref{fig:sroutliers} ), except for LLE, the number of highest density regions are not the same as the meta data. When comparing the number of HDRs for variable and fixed bandwidth (middle and right column in \autoref{fig:sroutliers} ), our proposed method with variable bandwidth outperforms fixed bandwith for ISOMAP, LLE, and Laplacian Eigenmaps (top three rows in \autoref{fig:sroutliers} ).
In terms of the top 20 outliers found rowwise, variable bandwidth could find most outliers lying on the left area of the embedding in ISOMAP and UMAP, and both methods in LLE embedding could find the outliers in the outer area, but for the other methods, both variable and fixed bandwidth are not detect true outliers accurately. For t-SNE and UMAP embedding, the embedding structure is highly distorted and the points are clustered together in a discontinuous way, which is also shown in the clustered outliers.

@@

To further compare the accuracy of the estimated densities for all data points, we calculate the correlation between the rank of true densities and the estimated densities and present in \autoref{tab:srcors}. It can be seen that the rank correlation of our proposed method with variable bandwidth is higher for LLE and UMAP, although the correlation for UMAP is very close to zero. The highest correlation comes from our method in LLE embedding, which is mainly due to it is closest to rectangular structure of the meta data shown in \autoref{fig:metadensity}. For Laplacian Eigenmaps, our method has wrongly estimated the left area with lower densities even though their true densities are the very high in yellow, leading to a negative correlation. The negative correlation would occur typically when the highest or lowest true density areas are not well estimated. As for the estimates in highly distorted embedding, including ISOMAP, t-SNE, and UMAP, the rank correlations are quite low. This shows that our proposed method could improve the kernel density estimate of manifold learning embedding by considering the distortion using the Riemannian metric. However, if the distortion is too severe, eg. ISOMAP, or when the embedding is discontinuous, eg. t-SNE and UMAP, the density estimates are not as reliable for outlier detection.

@@

We can gain further insight by comparing the correlation between ranks of true densities and estimated densities from varaible and fixed bandwidth KDE by \autoref{tab:tpcors}.
Again the highest correlations appears from embedding with higher quality, including ISOMAP, Laplacian Eigenmaps, and UMAP. The rank correlations between variable and fixed bandwidth is equivalent to the thrid decimal place in Laplacian Eigemaps and UMAP. As for t-SNE, the four clusters in the embedding are less seperated than in UMAP and our proposed method has misidentified the kernel cores, leading to a lower rank correlation in variable bandwidth. Since the embeddings from twin peak data generally capture the rectangular structure in the meta data than those from the swiss roll data, the rank correlations are much higher in \autoref{tab:tpcors} than in \autoref{tab:srcors}, with the lowest correlation being 0.259. This again suggests that the accuracy of outlier detection is highly related to the quality of manifold learning embedding.

@@

As a high-dimensional experiment, we generate the meta data from a 5-dimensional semi-hypersphere, transform it into a 100-dimensional space, and then embed it in \(d=5\) with manifold learning. First, we simulate \(N=2,000\) points, \((\pmb{X}_1, \pmb{X}_2, \pmb{X}_3, \pmb{X}_4)^\prime\), from a 4-dimensional Gaussian mixture model with two mixture components, \(\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)\) and \(\mathcal{N}(\pmb{\mu}_2, \pmb{\Sigma}_2)\), where \(\pmb{\mu}_1 = \pmb{\mu}_2 =(0, 0, 0, 0)^\prime\), \(\pmb{\Sigma}_1 = diag(1,1,1,1)\), and \(\pmb{\Sigma}_1 = diag(2,2,2,2)\). In order to manually add anomalies to be distant points from the means, the mixture proportions are set as \(\pi_1=0.99\) and \(\pi_2=0.01\).
The fifth dimension is calculated to satisfy the five-dimensional semi-hypersphere surface equation \(X_1^2 + X_2^2 + X_3^2 + X_4^2 + X_5^2 = r^2\) where \(X_5>0\) and \(r\) is set as \(7\).
The Gaussian mixture densities could be calculated using Equation \eqref{eq:gmm} as the true density of the 5-d meta data. \autoref{fig:fivedmeta} shows a scatterplot display when animating a 5-D tour path with the R package \emph{tourr} {[}REFERENCE{]}. The round and triangular point shapes index the two mixture components \(\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)\) and \(\mathcal{N}(\pmb{\mu}_2, \pmb{\sigma}_2)\), and the colors shows the distance between the simulated \(4-D\) data point from Gaussian mixture model and the kernel means \((0, 0, 0, 0)^\prime\). The more distant from the point to the kernel cores, the lower the true densities, which shows in a darker color in \autoref{fig:fivedmeta}. It can be seen that the most distant points are in triangular shape, meaning that they are simulated from \(\mathcal{N}(\pmb{\mu}_2, \pmb{\Sigma}_2)\). The dark colors also indicate that they are the true outliers because of their low densities.

>>>>

\begin{document}
\maketitle
\begin{abstract}
Kernel density estimation based on random samples often involves a crucial bandwidth selection. For data points lying on the Riemannian manifolds embedded in a high-dimensional space, manifold learning aims to find the low-dimensional representation of the data but there is local distortion for each observation, which failed to be captured by the kernel density estimation. We propose a new kernel density estimation of manifold learning embedding with the Riemannian metric being the pointwise bandwidth matrix and further identify outliers with the lowest densities. The Riemannian metric measures the direction and angle of the local distortion induced by the non-linear functions in manifold learning and is used for smoothing the density estimates. To compare our proposed method with fixed bandwidth KDE, we run two simulations with a 2-D meta data mapped into a 3-D swiss roll or twin peaks shape and a 5-D semi-hypersphere meta data mapped in 100-D space, and show that the proposed estimator could improve the density estimates given a good manifold learning embedding and gives higher correlations between ranks of true density and estimated density. For application, we use the Irish smart meter data and explore the distributions of different households and periods of the week. Both density estimators are compared by examing the distributions of the most typical households with highest densities and the most anomalous households with lowest densities.
\end{abstract}
\begin{keywords}
manifold learning, variable bandwidth, Riemannian metric, highest density region, Gaussian kernels
\end{keywords}

@@

Anomaly detection has been an important and diverse area where anomalies or outliers are detected in a given data. It often involves a data analysis process to uncover the unusual patterns and has been widely applied to machine learning \autocite{Omar2013-lz}, network intrusions identification \autocite{Ahmed2016-co,Bhuyan2013-ra}, medical imaging \autocite{Fernando2022-ef}, fraudulent attacks \autocite{Ahmed2016-dz}, cyber-security \autocite{Ten2011-co}, energy consumption \autocite{Cheng2021-ex}, the last of which explored detecting households with anomalous electricity usage distributions instead of raw data.
In the case of non-Euclidean sample space, the observations lie on a Riemannian manifold embedded in a very high-dimensional ambient space, which makes it computationally expensive or impossible to detect anomalies. To address this problem, we propose a kernel density estimator of the low-dimensional statistical manifold embedding and find outliers as the data with the lowest density estimates.

Kernel density estimation {[}KDE; \textcite{Parzen1962-gt}; \textcite{Chen2017-dw}{]} is one of the most popular methods to calculate the probability density function from a sample dataset. KDE is flexible to learn the shape of the underlying density from the data controlled by the bandwidth and the selection of bandwidth is crucial in KDE \autocite{Jones1990-oe,Terrell1992-ut}. There is extensive research on bandwidth selection, two main categories of which are cross-validation \autocite{Jones1992-ta,Sain1994-gr} and plug-in methods \autocite{Wand1994-gc,Duong2003-sp}. For multivariate data, variable kernel density estimator {[}VKDE; \textcite{Jones1990-oe}; \textcite{Terrell1992-ut}{]} has been studied with an adaptive bandwidth matrix to control the amount of smoothing on the location of the estimated point {[}balloon estimator; \textcite{Terrell1992-ef}{]} or the data point {[}sample smoothing estimator; \textcite{Terrell1992-ef}{]}.
However, these density estimators are based on random samples in the Euclidean space.

For samples points lying on a manifold with the differentiable structure called the Riemannian manifold, \textcite{Pelletier2005-vu} proposed a kernel density estimator based on the kernel weight depending on the distance between the estimated points and the sample data points. The idea of the estimator is to use a strictly positive function of the geodesic distance on the manifold and then normalize it with the volume density function of the Riemannian manifold for curvature. \autocite{Henry2009-ll} In many application scenarios, we tend to find that the sample points on a manifold are embedded in a much higher-dimensional space, and the kernel density estimator from \textcite{Pelletier2005-vu} is not directly applicable. This is when we introduce manifold learning to reduce the input data dimension.
For these high-dimensional data set, various manifold learning algorithms including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP (see details of these algorithms in \textcite{Cheng2021-ex}), are applied to get a low-dimensional embedding. We propose a kernel density estimator to be applied to the low-dimensional embedding and define outliers as those with the lowest densities.

In manifold learning, the underlying idea is that the data lies on a low-dimensional smooth manifold that is embedded in a high-dimensional space. One of the fundamental objectives of manifold learning is to explore the geometry of the dataset, including the distances between points and volumes of regions of data. These intrinsic geometric attributes of the data, such as distances, angles, and areas, however, can be distorted in the low-dimensional embedding, leading to failure to recover the geometry of the manifold \autocite{Goldberg2008-co}. To tackle this problem and measure the distortion incurred in manifold learning, \textcite{Perrault-Joncas2013-pq} propose the Learn metric algorithm to augment any existing embedding output with geometric information in the Riemannian metric of the manifold itself. By applying this algorithm, the outputs of different manifold learning methods can be unified and compared under the same framework, which would highly benefit in improving the effectiveness of the embedding.

The Riemannian metric using the method of \textcite{Perrault-Joncas2013-pq} gives some idea of the distortion of an embedding. Mapping the points through a non-linear function ``stretches'' some regions of space and ``shrinks'' others. The Riemannian gives us an idea of the direction and angle of this stretching, which is informative for learning the manifold.
In variable kernel density estimate, the bandwidth matrix \(H\) is also defined to control the amount of smoothing for each data point.
Therefore, we propose to the use Riemannian metric as the bandwidth matrix in, we could further get the kernel density estimation of the manifold \(\mathcal{M}\). This kernel density estimate can then be used to produce the highest density region plots \autocite{Hyndman1996-lk} for outlier visualization.

The rest of the paper is organized as follows.
In \autoref{vkde}, we present the proposed algorithm to detect anomalies based on variable kernel density estimates of manifold embeddings. In this section, we provide justification for the use of the Riemannian metric as the bandwidth of variable kernel density estimation, including the comparison with fixed bandwidth.
\autoref{simulation} is composed of two simulations with the proposed algorithm; the first deals with 2-dimensional meta data mapped into a 3-D swiss roll or twin peaks data and the second with a 5-D semi-hypersphere mapped in a 100-D space.
\autoref{application} contains the application to visualize and identify anomalies in the Irish smart meter dataset.
Conclusions and discussions are presented in \autoref{conclusion}.

@@

In this section, we introduce the proposed method to detect anomalies based on the kernel density estimates of manifold learning embeddings where the Riemannian matrix is used as the pointwise variable bandwidth to measure the direction and angle of the distortion of the low-dimensional embeddings. \textcite{Perrault-Joncas2013-pq} gives us an idea of how to measure the direction and angle of the distortion using the Riemannian metric and the Riemannian metric is a positive definite square matrix for each data point. To learn the distribution of the low-dimensional embedding, we use the kernel density estimation with the bandwidth matrix being the Riemannian metric. The outliers could then be defined as the points with the lowest density estimates. The proposed schematic is shown in Figure \ref{fig:vkde}. The highlighted two steps in Figure \ref{fig:vkde} are the main contributions of this main chapter, replacing the bandwidth matrix \(\pmb{h}_i\) with the Riemannian metric \(\pmb{g}_i\) for each point in variable kernel density estimate, and computing the highest density region plots based on the density estimates, \(\hat{f}(\pmb{x})\), for anomaly detection.

@@

In general, a multivariate variable kernel density estimate of \(f\) at point \(\pmb{x} \in \mathcal{R}^d\) is defined as
\[
\hat{f}(\pmb{x})=\sum\limits_{i=1}^N K_{\pmb{h}_i}(\pmb{x}-\pmb{a}_i),
\]
where \(a_i, i = 1, 2, \dots, N\) is the random sample drawn from a density \(f\), \(\pmb{h}_i\) is a \(d\times d\) symmetric positive definite matrix that is variable at \(i\), and \(K_{\pmb{h}_i}\) is the kernel function
\[
K_{\pmb{h}_i}(\pmb{x}-\pmb{a}_i)=(2\pi)^{-d/2}|\pmb{h}|^{-1/2}\exp\left[-\frac{1}{2}(\pmb{x}-\pmb{a}_i)'\pmb{h}_i^{-1}(\pmb{x}-\pmb{a}_i)\right],
\]
with \(|\pmb{h}_i|\) being the determinant of matrix \(\pmb{h}_i\). The bandwidth matrix is important in controlling the smoothing across all data points and the shape is determined by the kernel function \(K(\pmb{x})\). We use the sample smooth estimator for each data point, which makes a kernel density estimate with local smoothing.
\textcite{Duong2007-up} present an R package \emph{ks} that implement kernel density estimates with fixed bandwidth and variable bandwidth, but it is limited to bandwidth vector or diagonal matrix and does not apply to full bandwidth matrix.

@@

The Riemannian metric \(g\) is a symmetric and positive definite tensor field which defines an inner product \(<,>_g\) on the tangent space \(T_p\mathcal{M}\) for every point \(p \in \mathcal{M}\).
If the inner product of the tangent space is known for a given geometry, the Riemannian metric is a good measure to recover the geometry of the manifold.
The Metric Learning algorithm \autocite{Perrault-Joncas2013-pq} then augment the embedded manifold with the Riemannian metric and produce a Riemannian manifold \((\mathcal{M}, g)\).

@@

As pointed out by \textcite{Perrault-Joncas2013-pq}, if the embedding dimension \(s\) is larger than the manifold intrinsic dimension \(d\), the rank of the embedding metric \(h_n(p)\) is \(d\); otherwise, the Riemannian metric \(g_n\) will be returned.
This algorithm is also implemented in a Python library \emph{megaman} \autocite{McQueen2016-xz}. It is designed to apply the manifold learning methods to large-scale data sets, as well as compute the Riemannian metric of the manifold.

@@

If the embedding dimension \(d=2\), the HDR plots \autocite{Hyndman1996-lk} could be used to identify the relative location of outliers in the embedding and find which observations lie in the highest density region of specified coverage, eg. 1\%, 50\%, 99\%, \textgreater99\%.

@@

In this section, we examine two scenarios for both low and high dimensions to test our proposed algorithm. For visualization purposes, \autoref{twodgaussian} presents a 2-D meta data example. We first simulate the data of size \(N=2,000\) from a mixture of four Gaussian kernels with the same covariance but different means, each consisting of \(500\) points. Different mapping functions are then applied to the 2-D meta data to be mapped in a 3-D feature space, which gives the higher-dimensional input for different manifold learning algorithms, including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP. The embedded dimension is set as \(d=2\), the same as the meta data dimension. This enables us to compare the manifold learning embedding with the true meta data. We could now apply Algorithm \ref{alg:vkderm} to get the density estimates of all data points and further detect anomalies. As a high-dimensional example, the second simulation in \autoref{fivedgaussian} is based on a 5-D meta data of size \(N=2,000\) embedded in a 100-D space and the corresponding embedding dimension is \(d=5\).

@@

Given the 2-D meta data, multiple mapping functions could be applied to embed the data in a 3-D space. One of the most famous examples in manifold learning is the swiss roll data, with the mapping function in \eqref{eq:swissroll}. The two-dimensional meta data \((\pmb{X}_1, \pmb{X}_2)^\prime\) is transformed into the three-dimensional data \((\pmb{X}, \pmb{Y}, \pmb{Z})^\prime\), shown in the left plot of \autoref{fig:mappings}. The four colors in the mappings represent the four Gaussian kernels used to generate the meta data \((\pmb{X}_1, \pmb{X}_2)^\prime\).
\begin{equation}
\label{eq:swissroll}
\left\{
\begin{array}{lcl}
X = X_1 \cos{X_1}, \\
Y = X_2, \\
Z = X_1 \sin{X_1}.
\end{array}
\right.
\end{equation}

@@

\caption{Highest density region plots of five manifold learning embeddings of the swiss roll data. Colors are indicating densities from left: true densities from the Gaussian mixture model; middle: KDE with Riemannian matrix as variable bandwidth; and right: KDE with fixed bandwidth. Variable KDE performs better in finding kernel structures with ISOMAP, LLE, and Laplacian Eigenmaps, and in locating outliers with ISOMAP and LLE. The t-SNE and UMAP embeddings are highly distorted and the outliers found are clustered.}\label{fig:sroutliers}
\end{figure}

Now we are able to apply different manifold learning algorithms to \((\pmb{X}, \pmb{Y}, \pmb{Z})^\prime\) and reduce the dimension back to \(d=2\), and further estimate the density of the 2-D embedding. According to the density estimates, we could rank the data points and then identify which observations lie in the highest density region of specified coverage, eg. 1\%, 5\%, 50\%, 99\%, \textgreater99\%. For each of the five manifold learning methods, namely ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, \autoref{fig:sroutliers} presents the 2-D embedding plot in the same row, with the colors indicating the densities levels, the left column for true densities from the Gaussian mixture model, the middle column for highest density region plots with densities from our proposed variable KDE method, and the right for similar HDR plots with densities from KDE with fixed bandwidth. The top twenty outliers with the lowest densities are highlighted in black with point indexes in blue.
From \autoref{fig:metadensity} and the data generating process, we know that there are four highest density regions. However, in all manifold learning embeddings colored with true densities (left column in \autoref{fig:sroutliers} ), except for LLE, the number of highest density regions are not the same as the meta data. When comparing the number of HDRs for variable and fixed bandwidth (middle and right column in \autoref{fig:sroutliers} ), our proposed method with variable bandwidth outperforms fixed bandwidth for ISOMAP, LLE, and Laplacian Eigenmaps (top three rows in \autoref{fig:sroutliers} ).
In terms of the top 20 outliers found rowwise, variable bandwidth could find most outliers lying on the left area of the embedding in ISOMAP and UMAP, and both methods in LLE embedding could find the outliers in the outer area, but for the other methods, both variable and fixed bandwidth are not detected true outliers accurately. For t-SNE and UMAP embedding, the embedding structure is highly distorted and the points are clustered together in a discontinuous way, which is also shown in the clustered outliers.

@@

To further compare the accuracy of the estimated densities for all data points, we calculate the correlation between the rank of true densities and the estimated densities and present in \autoref{tab:srcors}. It can be seen that the rank correlation of our proposed method with variable bandwidth is higher for LLE and UMAP, although the correlation for UMAP is very close to zero. The highest correlation comes from our method in LLE embedding, which is mainly due to it being closest to the rectangular structure of the meta data shown in \autoref{fig:metadensity}. For Laplacian Eigenmaps, our method has wrongly estimated the left area with lower densities even though their true densities are very high in yellow, leading to a negative correlation. The negative correlation would occur typically when the highest or lowest true density areas are not well estimated. As for the estimates in highly distorted embedding, including ISOMAP, t-SNE, and UMAP, the rank correlations are quite low. This shows that our proposed method could improve the kernel density estimate of manifold learning embedding by considering the distortion using the Riemannian metric. However, if the distortion is too severe, eg. ISOMAP, or when the embedding is discontinuous, eg. t-SNE and UMAP, the density estimates are not as reliable for outlier detection.

@@

We can gain further insight by comparing the correlation between ranks of true densities and estimated densities from variable and fixed bandwidth KDE by \autoref{tab:tpcors}.
Again the highest correlations appear from embedding with higher quality, including ISOMAP, Laplacian Eigenmaps, and UMAP. The rank correlations between variable and fixed bandwidth are equivalent to the third decimal place in Laplacian Eigemaps and UMAP. As for t-SNE, the four clusters in the embedding are less separated than in UMAP and our proposed method has misidentified the kernel cores, leading to a lower rank correlation in variable bandwidth. Since the embeddings from twin peak data generally capture the rectangular structure in the meta data than those from the swiss roll data, the rank correlations are much higher in \autoref{tab:tpcors} than in \autoref{tab:srcors}, with the lowest correlation being 0.259. This again suggests that the accuracy of outlier detection is highly related to the quality of manifold learning embedding.

@@

As a high-dimensional experiment, we generate the meta data from a 5-dimensional semi-hypersphere, transform it into a 100-dimensional space, and then embed it in \(d=5\) with manifold learning. First, we simulate \(N=2,000\) points, \((\pmb{X}_1, \pmb{X}_2, \pmb{X}_3, \pmb{X}_4)^\prime\), from a 4-dimensional Gaussian mixture model with two mixture components, \(\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)\) and \(\mathcal{N}(\pmb{\mu}_2, \pmb{\Sigma}_2)\), where \(\pmb{\mu}_1 = \pmb{\mu}_2 =(0, 0, 0, 0)^\prime\), \(\pmb{\Sigma}_1 = diag(1,1,1,1)\), and \(\pmb{\Sigma}_1 = diag(2,2,2,2)\). In order to manually add anomalies to be distant points from the means, the mixture proportions are set as \(\pi_1=0.99\) and \(\pi_2=0.01\).
The fifth dimension is calculated to satisfy the five-dimensional semi-hypersphere surface equation \(X_1^2 + X_2^2 + X_3^2 + X_4^2 + X_5^2 = r^2\) where \(X_5>0\) and \(r\) is set as \(7\).
The Gaussian mixture densities could be calculated using Equation \eqref{eq:gmm} as the true density of the 5-d meta data. \autoref{fig:fivedmeta} shows a scatterplot display when animating a 5-D tour path with the R package \emph{tourr} {[}REFERENCE{]}. The round and triangular point shapes index the two mixture components \(\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)\) and \(\mathcal{N}(\pmb{\mu}_2, \pmb{\sigma}_2)\), and the colors shows the distance between the simulated \(4-D\) data point from Gaussian mixture model and the kernel means \((0, 0, 0, 0)^\prime\). The more distant from the point to the kernel cores, the lower the true densities, which shows in a darker color in \autoref{fig:fivedmeta}. It can be seen that the most distant points are in a triangular shape, meaning that they are simulated from \(\mathcal{N}(\pmb{\mu}_2, \pmb{\Sigma}_2)\). The dark colors also indicate that they are the true outliers because of their low densities.
